---
title: "DATS6101 Project 1"
author: "Mihir Gadgil, Elie Tetteh-Wayoe, Jessica Fogerty, Aluya Omofuma, Pierre Bamba"
date: "October 1, 2018"
output:
  html_document: default
  word_document: default
---


## Objectives 
#### First we imported the data from:   https://www.kaggle.com/heesoo37/120-years-of-olympic-history-athletes-and-results . 
#### Add Host_Country Column.
#### We noticed that each athlete was awarded a medal including each team member, we were able to subset the data so that each team was awarded a medal so that there was no discrepency.
#### Call the necessary packages for analysis. We used: ggplot2, dplyr.
#### Examine Structure of the data frame and then examined the five statistic summary.
#### Examine the formulated questions
#### Create visualizaions for analysis
#### Run appropriate tests

```{r setup, include = F}
library(ggplot2)
library(dplyr)
```

## EDA
Loading the dataset and reviewing its structure
```{r load_data, echo = F}
#Load the dataset
athletes <- read.csv("athlete_events.csv", na.strings = c("NA"))
#Explore the structure of the dataset
str(athletes)
```

We changed the structure of some of the columns such as medal, changed to ordered factor. 
We dropped Age, Height, Weight, City and Games. Select only the Summer Olympics data.
```{r cleanup_1, echo = F}
#athletes$Year <- as.integer(athletes$Year)
athletes$Medal <- factor(athletes$Medal, levels = c("Bronze", "Silver", "Gold"), ordered = TRUE)
athletes$NOC <- as.character(athletes$NOC)
#athletes$Gender <- as.factor(athletes$Gender)
athletes <- subset(athletes, select = -c(Age, Height, Weight, City, Games))
athletes <- subset(athletes, Season == "Summer")
```

Insert host country data into the data frame. Change some country codes to simplify analysis.
```{r cleanup_2, include = TRUE}
hosts <- data.frame(Year = c(seq(1896, 1912, 4), seq(1920, 1936, 4), seq(1948, 2016, 4)), Host_NOC = c("GRE", "FRA", "USA", "GBR", "SWE", "BEL", "FRA", "NED", "USA", "GER", "GBR", "FIN", "AUS", "ITA", "JPN", "MEX", "GER", "CAN", "RUS", "USA", "KOR", "ESP", "USA", "AUS", "GRE", "CHN", "GBR", "BRA"))
athletes <- merge(athletes, hosts, by.x = "Year", by.y = "Year")
athletes$Host_NOC[athletes$Year == 1956 & athletes$City == "Stockholm"] <- "SWE"
athletes$Host_NOC <- as.character(athletes$Host_NOC)
athletes$NOC[athletes$NOC == "URS"] <- "RUS"
athletes$NOC[athletes$NOC == "FRG"] <- "GER"
athletes$NOC[athletes$NOC == "GDR"] <- "GER"
athletes$NOC[athletes$NOC == "ANZ"] <- "AUS"
#Equating China and Hong Kong would help too, but it seems like they have been participating as different teams since 1952.
```
Interesting fact: 1906 had Olympic games, too. They were officially recognized as such at that time, but aren't considered Olympic games now.

Five summary statistics of the dataset
```{r after_cleanup, echo = F}
summary(athletes)
```
This summary shows us that the United States, Great Britain, and France are the teams that participated the most frequently. The sports that had the highest number of participants were: Athletics, Gymnastics, Swimming. It is important to note that the three types of medals are not equal in numbers. This is because of ties. 


## Analysis

### Does the Host Country have an advantage in Olympics?

First examine how many medals per country are won per olympic games
```{r all_medals, echo = F}
#Subset data to obtain only the medal winners 
winners_only <- subset(athletes, !is.na(Medal))
country_medals <- summarize(group_by(winners_only, Year, NOC), Medal_Count = n())
country_medals.plot <- ggplot(data = country_medals, aes(x = Year, y = Medal_Count, color = NOC)) + geom_point(show.legend = F) + labs(y = "Number of Medals", title = "Medals Earned by Countries over Years") + scale_x_continuous(breaks = seq(1896, 2016, 12)) + scale_y_continuous(breaks = seq(0, 450, 50))
country_medals.plot
```

```{r host_advantage, include = T}

#Find the total medal count per year
total_medals <- winners_only %>% group_by(Year) %>% summarize(Medal_Count = n())

#Include all of the host countries
hosts_only <- subset(winners_only, NOC %in% hosts$Host_NOC)

#Subset by medals won when the country that was a host country
medals_when_host <- subset(hosts_only, NOC == Host_NOC) %>% group_by(Year, NOC) %>% summarize(Medal_Count = n())
#Proportion of how many medals the host contry won out of all the medals avaliable for that year
host_medal_proportion <- data.frame(NOC = medals_when_host$NOC, Proportion = medals_when_host$Medal_Count / total_medals$Medal_Count)
#If a country hosted more than once, find the average of the proportion
host_medal_proportion <- host_medal_proportion %>% group_by(NOC) %>% summarize(Avg_Proportion = mean(Proportion))

#Subset by medals won when not acting as host country
medals_when_not_host <- subset(hosts_only, NOC != Host_NOC) %>% group_by(Year, NOC) %>% summarize(Medal_Count = n())
#Proportion of how many medals the host countries won out of medals avaliable when they are not host country
nonhost_medal_proportion <- merge(medals_when_not_host, total_medals, by.x = "Year", by.y = "Year")
nonhost_medal_proportion$Proportion <- nonhost_medal_proportion$Medal_Count.x / nonhost_medal_proportion$Medal_Count.y
#Average of proportion of medals won when a host country wasn't acting as a host country
nonhost_medal_proportion <- nonhost_medal_proportion %>% group_by(NOC) %>% summarize(Avg_Proportion = mean(Proportion))
```

Based on the visual inspection we would like to conduct a hypothesis test with the following hypothesis:  
$\mu_{D}$ is the difference between the average proportion of medals won as a host country and as a non host country  
$H_{0}: \mu_{D} = 0$  
$H_{1}: \mu_{D} > 0$  
Conduct a right tailed t-test to check whether a host country gets any advantage or not. Significance level $\alpha = 0.05$
```{r host_advantage2, include = T}
host_advantage_test <- t.test(host_medal_proportion$Avg_Proportion, nonhost_medal_proportion$Avg_Proportion, paired = T, conf.level = 0.95, alternative = "greater")
host_advantage_test
```
The p-value $`r host_advantage_test$p.value`$ is smaller than the significance level $0.05$.  
$\therefore$ We reject the null hypothesis.

# Participation of Men vs. Women
Take a look at the number of events conducted through years. Men/Women events counted separately.

```{r unique_events, echo = F}
unique_events_by_year <- data.frame(summarize(group_by(count(group_by(athletes, Year, Event)), Year), Events = n()))
unique_events_by_year.plot <- ggplot(data = unique_events_by_year, aes(x = Year, y = Events)) + geom_line(color = "red") + ylab("Number of Events") + labs(title = "Number of Events at each Olympic") + scale_x_continuous(breaks = seq(1896, 2016, 12)) + scale_y_continuous(breaks = seq(0, 325, 25))
unique_events_by_year.plot
```

Athlete participation over years.

```{r unique_participants, echo = F}

participation <- data.frame(summarize(group_by(count(group_by(athletes, Year, ID)), Year), Participants = n()))
participation.plot <- ggplot(data = participation, aes(x = Year, y = Participants)) + geom_line(color = "red") + labs(y = "Unique Athletes", title = "Unique Athletes at each Olympic") + scale_x_continuous(breaks = seq(1896, 2016, 12)) + scale_y_continuous(breaks = seq(0, 12000, 1500))
participation.plot
```

Men and Women Participation over the years.
```{r gender_participation, echo = F}
gender_participation <- data.frame(summarize(group_by(count(group_by(athletes, Year, Sex, ID)), Year, Sex), Participants = n()))
gender_participation <- add_row(gender_participation, Year = 1896, Sex = "F", Participants = 0, .before = 1)
gender_participation.plot <- ggplot(data = gender_participation, aes(x = Year, y = Participants, color = Sex)) + geom_line() + labs(title = "Number of Men and Women Participants over the years") + scale_x_continuous(breaks = seq(1896, 2016, 12)) + scale_y_continuous(breaks = seq(0, 7000, 1000))
gender_participation.plot
```

In fact the previous graph doesn't do the situation justice. The proportion of women participating each year gives a better picture. As of 2016 45% of the total participants were women. 
```{r women_participation, echo = F}
#Difference in medal count for men and women
women_participation <- data.frame(Year = hosts$Year, Proportion = gender_participation$Participants[gender_participation$Sex == "F"] / participation$Participants * 100)
women_participation.plot <- ggplot(data = women_participation, aes(x = Year, y = Proportion)) + geom_line(color = "red") + labs(y = "Percent of Total Participation", title = "Women's Participation in Olympics over the years") + scale_x_continuous(breaks = seq(1896, 2016, 12)) + scale_y_continuous(breaks = seq(0, 48, 8))
women_participation.plot
```

   
The linear model is:  
Proportion = Intercept + Slope $\times$ Year
```{r linear_model1, include = T}
#Create linear model for difference in medal count VS year
proportion_model <- lm(I(Proportion - 0) ~ I(Year - 1896) + 0, data = women_participation)
slope <- proportion_model$coefficients
```
```{r linear_model2, echo=F}
cat("Slope:", slope, end = "\n")
cat("Intercept:", 0)
```
The slope indicated the rate of increase of womens participation in percentage. Each year we see .3% more women (1.2% at each olympic games).

 
```{r linear_model3, echo = F}
prop_func <- function(year) {
  slope * (year - 1896)
}

women_participation.plot2 <- ggplot(data = women_participation, aes(x = Year, y = Proportion)) + stat_function(fun = prop_func) + labs(y = "Percent of Total Participation", title = "Women's Participation in Olympics over the years") + geom_point(color = "red") + scale_x_continuous(breaks = seq(1896, 2016, 12)) + scale_y_continuous(breaks = seq(0, 48, 8))
women_participation.plot2

cat("Predicted 50% year:", round(50 / slope + 1896))
```
By the year 2060 Men and Women participation will be equal.

### Cold war's effect on USA and Russia's performance
```{r before_cold_war, echo = F}
before_cold_war <- subset(athletes, Year < 1947 & (NOC == "USA" | NOC == "RUS"))
before_cold_war_participation <- data.frame(summarize(group_by(count(group_by(before_cold_war, Year, NOC, ID)), Year, NOC), Participants = n()))
before_cold_war_participation.plot <- ggplot(data = before_cold_war_participation, aes(x = Year, y = Participants, color = NOC)) + geom_line() + labs(title = "USA and Russia's participation before Cold War") + scale_x_continuous(breaks = seq(1896, 1936, 4)) + scale_y_continuous(breaks = seq(0, 550, 50))
before_cold_war_participation.plot
```

Select data that shows Russia and US performance before the cold war 
```{r cold_war, echo = F}
cold_war <- subset(athletes, Year >= 1947 & Year <= 1991 & (NOC == "USA" | NOC == "RUS"))
cold_war_participation <- data.frame(summarize(group_by(count(group_by(cold_war, Year, NOC, ID)), Year, NOC), Participants = n()))

cold_war_participation.plot <- ggplot(data = cold_war_participation, aes(x = Year, y = Participants, color = NOC)) + geom_line() + labs(title = "USA and Russia's participation during Cold War") + scale_x_continuous(breaks = seq(1948, 1988, 4)) + scale_y_continuous(breaks = seq(275, 550, 25))
cold_war_participation.plot
```



```{r cold_war2, echo = F}
russia_participation <- sum(subset(before_cold_war_participation, NOC == "RUS")$Participants)
usa_participation <- sum(subset(before_cold_war_participation, NOC == "USA")$Participants)
cat("Number of participants by Russia before the cold war", russia_participation, end = "\n")
cat("Number of participants by USA before the cold war", usa_participation)
russia_participation <- sum(subset(cold_war_participation, NOC == "RUS")$Participants)
usa_participation <- sum(subset(cold_war_participation, NOC == "USA")$Participants)
cat("Number of participants by Russia", russia_participation, end = "\n")
cat("Number of participants by USA", usa_participation)
```

Russia and US performance before the Cold War
```{r coldwar3, echo = F}
before_cold_war_performance <- summarize(group_by(subset(before_cold_war, !is.na(Medal)), Year, NOC), Winners = n())
before_cold_war_performance.plot <- ggplot(data = before_cold_war_performance) + geom_line(aes(x = Year, y = Winners, color = NOC)) + labs(y = "Number of Medals", title = "USA and Russia's Performance before the Cold War") + scale_x_continuous(breaks = seq(1896, 1936, 4)) + scale_y_continuous(breaks = seq(0, 400, 50))
before_cold_war_performance.plot
```

US outperformed Russia prior to the Cold War

Russia outperformed USA, in terms of medals won, in almost every Olympic during the Cold War.  
USA did not participate in the Olympics in 1980 and Russia did the same in 1948 and 1984.
```{r cold_war3, echo = F}
cold_war_performance <- summarize(group_by(subset(cold_war, !is.na(Medal)), Year, NOC), Winners = n())
cold_war_performance.plot <- ggplot(data = cold_war_performance) + geom_line(aes(x = Year, y = Winners, color = NOC)) + labs(y = "Number of Medals", title = "USA and Russia's Performance during the Cold War") + scale_x_continuous(breaks = seq(1948, 1988, 4)) + scale_y_continuous(breaks = seq(100, 450, 50))
cold_war_performance.plot
```

```{r cold_war4, echo = F}
russia_winners <- sum(subset(before_cold_war_performance, NOC == "RUS")$Winners)
usa_winners <- sum(subset(before_cold_war_performance, NOC == "USA")$Winners)
cat("Number of medals for Russia before the cold war:", russia_winners, end = "\n")
cat("Number of medals for USA before the cold war:", usa_winners, end = "\n")

russia_winners <- sum(subset(cold_war_performance, NOC == "RUS")$Winners)
usa_winners <- sum(subset(cold_war_performance, NOC == "USA")$Winners)
cat("Number of medals for Russia during the cold war:", russia_winners, end = "\n")
cat("Number of medals for USA during the cold war:", usa_winners)
```

Top 5 medal winners over all Olympics
```{r top_winners, include = TRUE}
country_medals.total <- country_medals %>% group_by(NOC) %>% summarize(Medal_Count = sum(Medal_Count))
country_medals.total <- data.frame(country_medals.total[order(country_medals.total$Medal_Count, decreasing = T),])
print(country_medals.total[1:5,])
```

Maximum medals earned by each sport, this tells us which sport is awarded the most medals by country. The USA is top in Swimming and Athletics 
```{r sport_winners, include = T}
country_medals.sport <- winners_only %>% group_by(NOC, Sport) %>% summarize(Medal_Count = n())

country_medals.sport <- country_medals.sport %>% group_by(Sport) %>% filter(Medal_Count == max(Medal_Count))
country_medals.sport <- country_medals.sport[order(country_medals.sport$Medal_Count, decreasing = T),]
print(country_medals.sport)
```

```{r echo = F}
sport_medals <- winners_only %>% group_by(Sport) %>% summarize(Medal_count = n())
sport_medals <- sport_medals[order(sport_medals$Medal_count, decreasing = T),]
print(sport_medals)
```

